---
title: "Genotype-by-Environment Interaction (GEI)"
author: "Igor Kuivjogi Fernandes"
date: "`r Sys.Date()`"
bibliography: references.bib
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

library(statgenGxE)
library(ggplot2)

data(dropsPheno)
dropsPheno$geno.panel <- NULL
dropsTD <- statgenSTA::createTD(data = dropsPheno, genotype = "Variety_ID", trial = "Experiment")
```

# Introduction

# Methodology

## Phenotypic Information

A data set from the European Union project called DROPS (DROught-tolerant yielding PlantS) was used in this study (<https://cordis.europa.eu/project/id/244374>). A panel of 256 maize hybrids was grown with two water regimes (irrigated or rainfed), in seven fields in 2012 and 2013, respectively, spread along a climatic transect from western to eastern Europe, plus one site in Chile in 2013. The panel resulted in 28 experiments defined as the combination of one year, one site and one water regime, with two and three repetitions for rainfed and irrigated treatments, respectively. A detailed environmental characterization was carried out, with hourly records of micro-meteorological data and soil water status, and associated with precise measurement of phenology. Grain yield and its components were measured at the end of the experiment. A total of 10 environments were selected from the full data set, two for each of the five main environmental scenarios (well watered and cool, well deficit and hot, well watered and hot, well deficit and hot in the day, and well watered and hot in the day), that were identified in the data, with 246 maize hybrids remaining. The trait analysed was the grain yield in tons per hectare (t/ha). Moreover, a classification of the genotypes in four genetic groups (Iodent, Lancaster, Stiff Stalk, and Other) was defined. The data is available at <https://doi.org/10.15454/IASSTN> [@millet2019].

## Models
### Mixed effects model
<!-- grain.yield ~ scenarioFull + scenarioFull:trial + (1 | genotype) + (1 | genotype:scenarioFull) -->
Multi-environment trial (MET) model was adjusted as follows:
\begin{align} \label{mod1}
  y = \mu + S + (S)t + \mathbf{G} + \mathbf{GS} + \mathbf{\varepsilon}
\end{align}
where $Y$ is the grain yield, $\mu$ if the intercept, $S$ is the fixed effect of the scenario, $(S)t$ is the fixed effect of trial nested within scenario, $\mathbf{G}$ is the random effect of genotype, $\mathbf{GS}$ is the random effect of the genotype-by-scenario interaction, and $\mathbf{\varepsilon}$ is the residual effect.

### Fully random effects model
A second model was adjusted, similar to model (\ref{mod1}), but with all terms as random effects to explore the different sources of variation. 

### Fully fixed effects model
A third model was performed, similar to model (\ref{mod1}), but with all terms as fixed effect, which consists of an Analysis of Variance (ANOVA) -- used to check the significance of treatments such as water availability and temperature (i.e. the $S$ effect).

### Finlay-Wilkinson Analysis
The Finlay-Wilkinson Analysis [@finlay1963] describes GEI by the heterogeneity of the slopes of a regression of individual genotypic performance on an environmental index. The environmental index is the average of all genotypes in an environment. The intercept expresses general performance across all environments, the slope represents adaptability, and the residuals may indicate a measure for stability.

The model fitted in the analysis was:
\begin{align} \label{finlay}
  y_{ij} = \mu + G_i + \beta_i t_j + \varepsilon_{ij}
\end{align}
where $y_{ij}$ is the grain yield of genotype $i$ in trial (environment) $j$, $\mu$ is the general mean, $G_i$ is the genotypic effect, $\beta_i$ a sensitivity parameter, $t_j$ the trial effect, and $\varepsilon_{ij}$ the residual term.

The model (\ref{finlay}) is fitted using an alternating regression algorithm. First, using starting values, $\beta_i$, $G_i$, and $t_j$ are estimated. Next, $t_j$ is assumed known and $\beta_i$ and $G_i$ are estimated. This process is continued until convergence, i.e. until the change in $\beta_i$ between iterations is less then a specified tolerance (e.g. 0.001). When estimating parameters, missing observations are estimated as well.

# Results

<!-- question 3: How much of the variance is explained by each variable? -->

The following table, calculated from the fully random effects model, shows the variance components and the variance explained by each variable:

```{r, warning = F, message = F}
dropsVarComp <- gxeVarComp(TD = dropsTD, trait = "grain.yield", nestingFactor = "scenarioFull")
varcomp <- dropsVarComp$fullRandVC
varcomp <- round(varcomp, 3)
varcomp$vcovPerc <- scales::label_percent()(varcomp$vcovPerc)
colnames(varcomp) <- c("variance", "% variance explained")
varcomp
```

<!-- question 1: What are the major genetic and nongenetic factors that are affecting the phenotypic variance? -->

The major nongenetic effect is `scenarioFull`, which is the environmental scenario, and explains 75.5% of the total phenotypic variability. The major genetic effect is `genotype`, and explains 11.6% of the total phenotypic variability.

An ANOVA table was also provided from the fully fixed effects model:

```{r}
dropsVarComp$aovFullFixedMod
```

<!-- question 2: What is the effect of water availability and temperature on the variation of grain yield? -->

The water availability and temperature (the term `scenarioFull`) was significant at an significance level of $\alpha = 0.05$, thus grain yield was affected by the effect of water availability and temperature.

<!-- question 4: What is the heritability of grain yield in this study? -->

From the variance components table of the fully random effects model, to analyse how much of the phenotypic variability is due to genetic effect, the heritability was calculated based on @atlin2000, and showed a value of `r herit(dropsVarComp)`.


<!-- question 5: Which environments have the highest and lowest environmental effects? -->

From the Finlay-Wilkinson Analysis it is possible to check the environmental effects per trial. For example, `Gai12W` showed the highest environmental effect (4.291), whereas `Cra12R` had the lowest environmental effect (-5.417). The other environmental effects can be seen in the following table: 
```{r}
dropsFW <- gxeFw(TD = dropsTD, trait = "grain.yield")
dropsFW$envEffs[order(dropsFW$envEffs$EnvEff, decreasing = T), ]
```

<!-- question 6: Which genotypes are the most sensitive and what does this metric represent? Include a plot comparing the genotype with the highest and lowest sensitivity. -->
In addition to analysing environmental effects, it is also possible to explore the interaction of genotypes and environments. In the Finlay-Wilkinson model (\ref{finlay}), the slopes $\beta_i$ are a measure for the sensitivity of genotype $i$ to the improvement in the quality of the environment (also called adaptability), where there is one slope per genotype (in this case 246 slopes). In other words, adaptability is the ability of a genotype to respond positively to changes in environments @chloupek2005. The following table shows the least and the most sensitive genotypes:
```{r}
sens <- dropsFW$estimates[order(dropsFW$estimates$Sens), ]
sens[c(1, nrow(sens)), ]
```
The genotype `F252` was the least sensitive, whereas `Lo1251` was the most sensitive. In other words, `F252` was the genotype that less improves the environment quality, while `Lo1251` was the genotype that most improves the environment quality. This phenomenon can be seen through a "slope plot":  

```{r, include = F}
p1 <- plot(dropsFW, plotType = "trellis", genotypes = c("F252", "Lo1251"),
     title = "Less and most sensitive genotypes")
```

```{r, fig.width = 5, fig.height = 3}
p1 + labs(x = "Environmental mean") 
```


In the plot above, the genotype `Lo1251` has a larger slope (larger angle to the x-axis) than the less sensitive genotype. The x-axis represents the environmental mean, and there are 10 points, one point per trial. As the environmental mean increases, the grain yield for genotype `Lo1251` increases in a larger extent when compared to the genotype `F252`. 

It is interesting to note that genotypes with slopes > 1 have higher than average sensitivity, and genotypes with slopes < 1 have lower than average sensitivity because the mean of slopes is 1, i.e., $\frac{1}{n} \sum_{i=1}^{n} \beta_i = 1$, $i = 1, ..., 246$.

## References
