---
title: "Genotype by Environment (GxE) interaction"
author: "Igor Kuivjogi Fernandes"
date: "`r Sys.Date()`"
bibliography: references.bib
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

library(statgenGxE)

data(dropsPheno)
dropsPheno$geno.panel <- NULL
dropsTD <- statgenSTA::createTD(data = dropsPheno, genotype = "Variety_ID", trial = "Experiment")
```


## Introduction

## Methodology
### Phenotypic Information
A data set from the European Union project called DROPS (DROught-tolerant yielding PlantS) was used in this study (https://cordis.europa.eu/project/id/244374). A panel of 256 maize hybrids was grown with two water regimes (irrigated or rainfed), in seven fields in 2012 and 2013, respectively, spread along a climatic transect from western to eastern Europe, plus one site in Chile in 2013. The panel resulted in 28 experiments defined as the combination of one year, one site and one water regime, with two and three repetitions for rainfed and irrigated treatments, respectively. A detailed environmental characterization was carried out, with hourly records of micro-meteorological data and soil water status, and associated with precise measurement of phenology. Grain yield and its components were measured at the end of the experiment. A total of 10 environments were selected from the full data set, two for each of the five main environmental scenarios (well watered and cool, well deficit and hot, well watered and hot, well deficit and hot in the day, and well watered and hot in the day), that were identified in the data, with 246 maize hybrids remaining. The trait analysed was the grain yield in tons per hectare (t/ha). Moreover, a classification of the genotypes in four genetic groups (Iodent, Lancaster, Stiff Stalk, and Other) was defined. The data is available at https://doi.org/10.15454/IASSTN [@millet2019].

### Genotype-by-environment analysis
Multi-environment trial (MET) analysis were used to study Genotype-by-environment interaction (GEI). The following mixed linear model was adjusted:
```
% grain.yield ~ scenarioFull + scenarioFull:trial + (1 | genotype) + (1 | genotype:scenarioFull)
Y = S + (S)t + \mathbb{g} + \mathbb{(g)S} + \mathbb{e},
```
where $Y$ is the grain yield, $S$ is the fixed effect of the scenario, $(S)t$ is the effect of trial nested within scenario, $\mathbf{g}$ is the random effect of genotype, $\mathbf{(g)S}$ is the scenario nested within genotype, and $\mathbf{e}$ is the residual effect.


## Results

```{r}
# traits relationships
# pairs(dropsPheno[, c("grain.yield", "grain.number", "seed.size", "anthesis", "silking", "plant.height", "tassel.height", "ear.height")])
```


```{r}
# plot(dropsTD, plotType = "box", traits = "grain.yield", colorTrialBy = "scenarioFull",
#      orderBy = "descending")
```

The following table shows the variance components: 
```{r}
# dropsVarComp <- gxeVarComp(TD = dropsTD, trait = "grain.yield", nestingFactor = "scenarioFull")
varcomp <- dropsVarComp$fullRandVC
colnames(varcomp) <- c("variance", "% variance explained")
varcomp <- round(varcomp, 3)
varcomp
```

The nongenetic effects are:
- `scenarioFull`, which is the environmental scenario, explains 75.5% of the total variability
- ``

```{r}
# diagnostics(dropsVarComp)
```

```{r}
vc(dropsVarComp)
```

```{r}
herit(dropsVarComp)
```

## References

